# Introduction

Lung cancer is one of the most common types of cancer in both men and women. The analysis of survival time of patients with lung cancer is crucial for controlling the disease's development, determining the optimal course of treatment, and understanding what influences the disease's progression. To accomplish these aims, accurate survival time analysis methods are crucial.

Survival analysis is the combination of different statistical methods for analysing time-to-event data. Executing a survival analysis can be challenging, since both non-parametric and parametric models can be used, and various distributions such as exponential, Weibull, and log-normal are often all applicable. The most common model is the Cox hazard model which proposes a constant effect of predictor variables on survival duration (known as the hazard function) throughout time, which we know to be unrealistic, and thus consider to be too simple for this case.

This study employs a Bayesian Weibull Generalised Linear model (GLM) of the survival time of patients with advanced lung cancer. The Weibull approach is more flexible than Cox, and the hazard rate is not assumed to be constant with time. For simulation Bayesian inference with Markov Chain Monte Carlo (MCMC) is used, providing us with a satisfying approximation of uncertainty and ability to use priors from domain knowledge. The model is implemented and tested with the help of the R and Stan packages.

We will investigate the efficacy of three models: a model with data pooled across insitutions and trained on only observed data, a similarly pooled model using censored data, and a hierarchical model using only observed data. The respective models will eventually be compared with regards to their predictive power, and we will conclude that our hierarchical model is the most performant and conclude with a discussion on its implications.

The code for the report, complete with model implementations in R and Stan, is provided in @git.

## Data description

### General description

The data used in the study include the survival time of patients with advanced lung cancer from the North Central Cancer Treatment Group. It is provided in the `survival` R package from @survival-package. We will aim to predict the survival time of patients with lung cancer given some covariates.

Our dataset contains 9 features and 149 observations, which we divide into 7 institutional groups, and which we assume to be independent and identically distributed. The target variable is the survival time in days. The covariates are of both categorical and numerical data types.

Special attention has to be paid to the "censored status" indicator. It indicates if the patient had some terminal event (in which case it is equal to $1$) or not (equal to $0$). If the data for a patient are censored, their true survival time is not known. A right censoring approach is used, meaning incompleteness of survival time at the right side of the follow-up period.

Our variables, and their descriptions adapted from @survival-package, are as follows:

Variable | Description
---|---
`inst`|	Institution code
`time`|	Survival time in days
`status`|	censoring status 1=censored, 2=dead
`age`|	Age in years
`sex`|	Male=1 Female=2
`ph.ecog`|	ECOG performance score (0-4). 0-good condition, 4-the worst condition.
`ph.karno`|	Karnofsky performance score (bad=0-good=100). Provided by physician.
`pat.karno`|	KKarnofsky performance score. Provided by patient.
`meal.cal`|	Calories consumed at meals
`wt.loss`|	Weight loss in last six months (pounds)

### Exploratory data analysis

We provide some descriptive statistics to familiarise ourselves with the data, beginning by importing the necessary packages.

```{r,warning=FALSE,message=FALSE,results=FALSE}
library(dplyr)
library(ggplot2)
data("cancer", package = "survival")
```

There were 44 observations with missed values, which have been removed from dataset, leaving 105 observations.

```{r}
data = cancer %>% 
  filter(inst %in% c(1, 12, 13, 3, 11, 22, 16)) %>%
  na.omit()
```

Institutions are considered as grouping variables, used later in the hierarchical model. Note that to make this analysis more manageable, we are only considering data from the top 7 institutions in terms of number of observations for all models.

```{r}
table(data$inst)
```

Moving to categorical variables below in Figure 1, we have more men than women in our data, and note that the majority of patients are ambulatory with symptoms.

```{r,fig.width=8,fig.height=4,fig.cap="Categorical variables"}
par(mfrow=c(1,2))
barplot(table(data$sex), main="Sex statistics", names.arg=c("male", "female"),
        col=c("steelblue","cornflowerblue"))
barplot(table(data$ph.ecog), main="ECOG score statistics",
        legend = c("asymptom.", "ambulatory", "in bed <50% of t", "in bed >50% of t"),
        args.legend = list(x = "topright",inset = c(- 0.15, 0)),
        col=c("steelblue","cornflowerblue","blue","darkblue"))
```

Investigating the distribution of continuous variables in Figure 2, we find that the features do not follow a normal distribution.

```{r,fig.width=6,fig.height=6,fig.cap="Continuous variables"}
par(mfrow=c(3,2))
hist(data$age, freq=FALSE, col="cornflowerblue", main="Histogram of age",xlab="")
hist(data$ph.karno, freq=FALSE, col="cornflowerblue", main="Histogram of ph.karno",xlab="")
hist(data$pat.karno, freq=FALSE, col="cornflowerblue", main="Histogram of pat.karno",xlab="")
hist(data$meal.cal, freq=FALSE, col="cornflowerblue", main="Histogram of meal.cal",xlab="")
hist(data$wt.loss, freq=FALSE, col="cornflowerblue", main="Histogram of wt.loss",xlab="")
```

It is also useful to identify any linear correlations between variables. The correlation is not that high across the board, and only the pair `ph.karno` and `pat.karno` are highly correlated (0.525), but this is intuitively reasonable given the previous description of these covariates, since patients and doctors may measure the same quantity, or agree on the quantity measured.

```{r}
cor(data[c(4,7,8,9,10)], method=c("pearson"))
```

## Related studies

The original data were presented in @original, where the authors just provided descriptive information from a lung patient-completed questionnaire, which was aggregated into the given dataset. In @parametric-survival the comparison of semi-parametric and non-parametric models for survival analysis was presented, but a different dataset was used, and there was no serious description of Weibull model implementation. Instead, more attention was payed to Cox regression. @exponential-model provided a description of exponential models applied to a lung cancer survival time analysis with Stan code, but the Weibull distribution was only mentioned in passing and no models or conclusions were derived for it. That is why in our work three Weibull Survival models are built: one hierarchical and two pooled which will all be described in the following section. The main approaches and survival analysis theory were inspired from the studies above, but the Stan implementation, data preprocessing, choice of priors, and interpretation were all performed by the authors.
