# Description of models

In the following section, we will motivate and define mathematically our three GLMs implemented in Stan for survival time prediction. We begin by importing the necessary packages and setting a random seed for reproducibility.

```{r, message=FALSE, warning=FALSE}
# install libraries
library(survival)
library(tidyverse)
library(rstan)
library(bayesplot)
library(loo)
library(ggplot2)
library(data.table)
# set number of cores
options(mc.cores = parallel::detectCores())
# read lung cancer data from `survival` library
data("cancer", package = "survival")
# set random seed for reproducibility
set.seed(2021)
```

## Weibull without censored data

Our first model will be a pooled model, not considering any censored data points. Let $y \sim \text{Weibull}(\alpha, \sigma)$, we define the probability density function or our Weibull-distributed $y$ as
$$
\text{Weibull}(y|\alpha,\sigma) =
\frac{\alpha}{\sigma} \, \left( \frac{y}{\sigma} \right)^{\alpha - 1}
\, \exp \! \left( \! - \left( \frac{y}{\sigma} \right)^{\alpha}
\right),
$$
for $y \in [0,\infty), \alpha \in \mathbb{R}^+,$ and $\sigma \in \mathbb{R}^+$. 

### Motivating the distribution

The Weibull distribution is often used as a more flexible and complex alternative to the semi-parametric proportional hazard Cox model for modelling time to failure events, since the hazard rate is not taken to be constant with time. We will use this as our base for all models in this analysis to differentiate ourselves from previous analyses, and hopefully outperform them.

### The Weibull distribution as a member of the Exponential family

A probability distribution $f(y\vert\vartheta)$ is a member of the Exponential family if it can be written in the form
$$
f(y\vert\vartheta) = h(y)\exp[\eta(\vartheta) \cdot T(y) - A(\vartheta)],
$$
for some arbitrary parameter functions $h(\cdot), T(\cdot), A(\cdot)$, and canonical parameter $\eta(\vartheta)$ as a function of the parameters $\vartheta = (\sigma, \alpha)^T$. Now take $\alpha$ fixed and finite, then it can be shown that the Weibull distribution belongs to the exponential family since we can write it's probability density function as
$$
\text{Weibull}(y|\sigma) ={\alpha y^{\alpha -1}}\exp(-y^{\alpha}\sigma^{-\alpha}-\alpha\log\sigma),
$$
with
\begin{align*}
b(y)&={\alpha y^{\alpha -1}} \\
\eta(\vartheta)&=\sigma^{-\alpha}\\
T(y)&=-y^{\alpha} \\
a(\eta)&=\alpha\log\sigma.
\end{align*}
This will make inference in the future easier for us.

### Defining the link function

Looking at our canonical statistic $\eta=\sigma^{-\alpha}$, it can be shown that
$$
\sigma=\exp{({\frac{\log\eta}{-\alpha}})}
$$
where we construct $\eta = \exp(\boldsymbol X\beta)$ so that $\eta$ is always strictly positive given any linear combination of the covariates, and the logarithm is then defined. Thus we choose a log link function for our GLM such that
$$
\sigma = \exp(-\frac{\boldsymbol X\beta}{\alpha}),
$$
which resembles a vanilla logarithmic inverse link function, with the additional $\alpha$ term and the negation.

### Priors

In our Stan model, we will enforce priors over each of the regressors in the linear model, $\beta$, and the shape parameter $\alpha$ of our Weibull distribution. Mathematically, where we have $N$ data points and $M$ covariates, the model is defined as
\begin{align*}
y_i &\sim \text{Weibull}(\sigma, \alpha)\,,\quad i = 1,\dots,N,\\
\alpha &\sim \text{Half-Cauchy}(5),\\
\sigma &= \exp\left ( -\frac{\boldsymbol X \beta}{\alpha}\right ),\\
\beta_j &\sim N(0, 1)\,,\quad j=1,\dots,M.\\
\end{align*}
The choice of a Half-Cauchy prior is motivated by @taumain, so that the posterior is more flexible than the Gamma alternative, and the data can express themselves more freely. Note that this is a specific case of the of the conditionally-conjugate folded-noncentral-t family of prior distributions. In short, it acts as an appropriate weakly-informative prior satisfying the positive constraint, and given that we expect the value of $\alpha$ to be below 10 but allow for some deviance from this range. Our Gaussian prior over the regressors is easily motivated by the parameterisation of the data, and is standard in literature. We will mean-center the design matrix before performing our regression in a `transformed data` block. This standardisation will hopefully reduce variance in our regressors and make learning their weights easier. We expect the weights of the regressors to be close to zero, and we allow them to be negative, making this an appropriate prior. In fact, the fact that the variance is not very large should aid convergence later, since our priors are stronger than for example a $N(0, 100)$ prior.

In this pooled model, we model these parameters using data from all institutions. Intuitively, this means that we expect the hazard to be equivalent regardless of which institution a patient is in. This is seen visually in the directed acyclic graph in Figure 3 below.

```{r, echo=FALSE, out.width="50%", fig.align="center", fig.cap="Pooled model DAG"}
knitr::include_graphics("./rmd/pooled.pdf")
```

### Implemented in Stan

Below, we fit the model in Stan and output its code for the reader. Convergence and other diagnostics will be presented in the following section for all models.

```{r weibull, warning=FALSE, message=FALSE}
# build dataset from only those non-censored data points
uncensored_data <- cancer %>%
  filter(status == 2) %>%
  filter(inst %in% c(1, 12, 13, 3, 11, 22, 16)) %>%
  drop_na()
# identify covariate labels and build design matrix
cov_labels <- uncensored_data %>%
  dplyr::select(-status,-time,-inst) %>%
  colnames()
# build design matrix
X <- as.matrix(uncensored_data[cov_labels])
# print(dim(X))
# [1] 120   7
y <- uncensored_data$time
# build data list for Stan model
weibull_data = list(
  y = y, X = X, N = length(y), M = ncol(X)
)
# compile and run seperate model
wm <- rstan::stan_model(file = "../stan/weibull_survival.stan")
# print out Stan code
print(wm)
# learn the model parameters
weibull_model <- rstan::sampling(wm, iter = 10000, data = weibull_data)
```

## Weibull with censored data

The density function for Weibull distributed survival times is given as

$$
{p(t_{i}|\alpha,\lambda_{i}) = \alpha t_{i}^{\alpha-1}\exp{(\lambda_{i}-\exp{\lambda_{i}}t_{i}^{\alpha})}},
$$

and can be rewritten as

$$
{p(t_{i}|\alpha,\gamma_{i}) = \exp{\bigg(-\bigg(\frac{t_{i}}{\gamma_{i}}\bigg)^{\alpha}\bigg)}\frac{\alpha}{\gamma_{i}}\bigg(\frac{t_{i}}{\gamma_{i}}\bigg)^{\alpha-1}},
$$
where $\alpha$ is the shape parameter, and $\gamma$ the scale. We move on to define a new variable $\lambda$ is created, defined in relation to $\gamma$ as

$$
{\lambda=-\alpha \log{\gamma}}.
$$

The survival function, showing the probability that the death will be after a certain time t, is then

$$
{S(t_{i}|\alpha,\lambda_{i})=\exp(-\exp(\lambda_{i})t_{i}^{\alpha})}.
$$

The likelihood of $\alpha$ and $\lambda$ follows the equation below, with $v_{i}$ an indicator showing 0 if the data are censored and 1 if not,

$$
{L(\alpha,\lambda|t)=\prod\limits_{i = 1}^n p(t_{i}|\alpha,\lambda_{i})^{v_{i}} S(t_{i}|\alpha,\lambda_{i})^{1-v_{i}}}=
\prod\limits_{i = 1}^n (\alpha t_{i}^{\alpha-1}exp(\lambda_{i}))^{v_{i}}(exp(-exp(\lambda_{i})t_{i}^{\alpha})).
$$

If $\lambda=X\beta$, than log-likelihood function can be expressed as,

$$
l(\alpha,\beta|t,x)=\sum\limits_{i = 1}^nv_{i}(\log(\alpha)+(\alpha-1)log(t_{i})+X_{i}\beta)-\exp(X_{i}\beta)t_{i}^{\alpha}.
$$
If the data are censored, then the log-likelihood is simply the logarithm of the survival function. In Stan this can be expressed with `weibull_lccdf()` function, corresponding to the log of the Weibull complementary cumulative distribution function of $y$ given shape $\alpha$ and scale $\sigma$, and it is exactly the logarithm of survival function. For clarity, the complementary cumulative distribution function is
$$
\bar{F}_{X}(x)=P(X>x)=1-F_{X}(x),
$$
where $F_{X}(x)$ is the cumulative distribution function of a random variable.

### Priors

The same priors are used for the censored model as for the pooled model. Even we are not expert oncologists, we know that typically person's life with advanced lung cancer is a little bit lower than a year, some patients live even for three years. The chosen prior is suitable, as it allows to produce survival time values, which are not too strict and at the same time really unlike to be larger than 5 years, for example.

Chosen priors do not contribute strongly to the posterior, so the data can express themselves. Algebraically then, the model is given as
\begin{align*}
y_i &\sim \begin{cases}\text{Weibull}(\sigma, \alpha)\,,\quad \text{if }y_i\text{ is observed},\\
\text{log-likelihood Weibull}(\sigma, \alpha)\,,\quad \text{if }y_i\text{ is censored},
\end{cases}\\
\alpha &\sim \text{Half-Cauchy}(5),\\
\sigma &= \exp\left ( -\frac{\boldsymbol X \beta}{\alpha}\right ),\\
\beta_j &\sim N(0, 1)\,.\\
\end{align*}
Note that this is graphically equivalent to the pooled model.

### Implemented in Stan

We once more execute some data preprocessing.

```{r, warning=FALSE, message=FALSE}
# read lung cancer data from "survival' library
data("cancer", package = "survival")

# omittimg NAs
data = cancer %>% 
  filter(inst %in% c(1, 12, 13, 3, 11, 22, 16)) %>%
  na.omit()

# Censoring status is transformed to the column with 0-censored, 1-observed.
# The continuous variables are centered.
X=data
X$status[X$status==1] = 0
X$status[X$status==2] = 1
#X$male = ifelse(X$sex==1,1,0)
#X$female = ifelse(X$sex==2,1,0)
X$age=X$age-mean(X$age)
X$meal.cal=X$meal.cal-mean(X$meal.cal)
X$wt.loss=X$wt.loss-mean(X$wt.loss)

Xcens=X[X$status==0,]
Xcens = Xcens[-c(1,2,3)]
ycens=X$time[X$status==0]

Xobs=X[X$status==1,]
Xobs = as.matrix(Xobs[-c(1,2,3)])
yobs=X$time[X$status==1]
```

And collate the data in such a way as to present them to the Stan model.

```{r, warning=FALSE, message=FALSE, comment=FALSE}
data_model = list(
  yobs = yobs,
  Xobs = Xobs,
  N = nrow(Xobs),
  M = ncol(Xobs),
  ycen = ycens,
  Xcen = Xcens,
  Ncen = nrow(Xcens)
)
```

We are now able to run the model and display the Stan code. 

```{r censored, warning=FALSE, message=FALSE}
# compile and run censored model
cwm = rstan::stan_model(file = "../stan/weibull_censored.stan")
# print out Stan code
print(cwm)
# learn the model with parameters 4 chains, 10000 iterations for each, 5000 iterations for warm-up
weibull_cens = rstan::sampling(cwm, data = data_model, iter = 10000)
```

## Hierarchical Weibull without censored data

Here we implement a model with some global shape parameter $\alpha$ to be learned as in the pooled model, but independent regressor parameters $\beta_j$ for each institution $j$. Once more, we ignore the censored data. By virtue of this, we need not worry about the more complex distribution functions shown above, but we do sacrifice the number of data points we can learn from for each institution. We now consider our model in terms of the same $N$ data points and $M$ regressors, but we will estimate the covariates' weights according to the institution, of which we have $J$ in total. Thus our model is defined mathematically as
\begin{align*}
y_{ij} &\sim \text{Weibull}(\sigma_j, \alpha)\,,\quad i = 1,\dots,N,\,j= 1,\dots,J\\
\alpha &\sim \text{Half-Cauchy}(5),\\
\sigma &\propto \boldsymbol X \beta,\,\text{via previously seen link function}\\
\beta_{mj} &\sim N(\mu_\beta, \sigma_\beta)\,,\quad m=1,\dots,M,\\
\mu_\beta &\sim N(0, 1),\\
\sigma_\beta &\sim \text{Gamma}(1, 1).
\end{align*}
This is seen visually in the directed acyclic graph in Figure 4 below.

```{r, echo=FALSE, out.width="50%", fig.align="center", fig.cap="Hierarchical model DAG"}
knitr::include_graphics("./rmd/hier.pdf")
```

### Defining the link function

The link function used is the same as that shown in the first model, and is similarly motivated.

### Priors

The same global priors are used for the hierarchical model as the pooled model. 

We motivate a Gamma hyperprior over the variance of $\beta$ with @taumain, since we have 7 different groups, a Half-Cauchy will be too sensitive to different group data whereas a Gamma will be more stable for more groups and aid convergence as a result. We have a standard Gaussian hyperprior over the the mean of $\beta$, which we can motivate since the value of $\mu_\beta$ can be either positive or negative so this allows the groups to express themselves either way, and this is a standard hyperprior over a mean we assume to have zero mean.

### Implemented in Stan

We once more engage in data preparation and run the Stan model, displaying the code, before moving on to the next section where we will discuss the diagnostics of each model.

```{r hier_weibull, warning=FALSE, message=FALSE}
# build dataset from only those non-censored data points
uncensored_data <- cancer %>%
  filter(status == 2) %>%
  filter(inst %in% c(1, 12, 13, 3, 11, 22, 16)) %>%
  drop_na()
# identify covariate labels and build design matrix
cov_labels <- uncensored_data %>%
  dplyr::select(-status, -time, -inst) %>%
  colnames()
# build design matrix
X <- as.matrix(uncensored_data[cov_labels])
# observed survival times
y <- uncensored_data$time
# institution labels
ll <- as.numeric(as.factor(uncensored_data$inst))
# build some hierarchical data for Stan
hier_data = list(
  y = y,
  X = X,
  ll = ll,
  N = length(y),
  M = ncol(X),
  J = length(unique(ll))
)
# compile and run seperate model
whm <- rstan::stan_model(file = "../stan/weibull_hier.stan")
# print out Stan code
print(whm)
# learn the model parameters
weibull_hier <- rstan::sampling(whm, data = hier_data, iter = 10000)
```

